\documentclass[11pt]{article}

\usepackage[margin=0.9in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}

\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
\definecolor{codeblue}{rgb}{0.05,0.15,0.55}
\definecolor{codegreen}{rgb}{0.0,0.45,0.0}

\lstset{
    backgroundcolor=\color{backcolour},
    basicstyle=\ttfamily\small,
    keywordstyle=\color{codeblue}\bfseries,
    commentstyle=\color{codegreen},
    stringstyle=\color{purple},
    breaklines=true,
    frame=single,
    showstringspaces=false,
    tabsize=4
}

\title{CPSC 250 \\ Advanced Linear Regression Concepts}
\author{Edward Brash}
\date{}

\begin{document}

\maketitle

\section{Motivation}

In earlier lectures, we considered simple linear regression involving one independent variable.

Real-world data analysis is often more complicated.

In this lecture, we discuss:

\begin{itemize}
    \item multiple regression variables,
    \item parameter interpretation,
    \item correlations between variables,
    \item matrix representations,
    \item and data structures for regression.
\end{itemize}

\section{Complex Regression Problems}

Suppose we want to predict:

\[
y
\]

using several variables:

\[
x_1, x_2, x_3, x_4
\]

For example:

\begin{itemize}
    \item \(y\): housing price,
    \item \(x_1\): square footage,
    \item \(x_2\): zip code,
    \item \(x_3\): number of bedrooms,
    \item \(x_4\): number of bathrooms.
\end{itemize}

\section{The Regression Equation}

We seek an equation of the form:

\[
y =
\beta_0
+
\beta_1 x_1
+
\beta_2 x_2
+
\beta_3 x_3
+
\beta_4 x_4
\]

This is linear in the parameters.

\section{Why This Is Difficult}

Visualizing this problem directly would require a very high-dimensional plot.

For example:

\begin{itemize}
    \item one dependent variable,
    \item four independent variables.
\end{itemize}

This is impossible to visualize completely.

Therefore, regression becomes both:

\begin{itemize}
    \item a computational problem,
    \item and an interpretation problem.
\end{itemize}

\section{Parameter Uncertainties}

In practice, we determine:

\[
\beta_0 \pm \Delta\beta_0
\]

\[
\beta_1 \pm \Delta\beta_1
\]

and so on.

A parameter consistent with zero may indicate that the corresponding variable is not important.

For example:

\begin{quote}
If the coefficient for number of bedrooms is consistent with zero, housing prices may not strongly depend on that variable.
\end{quote}

\section{Exploratory Data Analysis}

Before fitting, we should first examine the data.

For example:

\[
y \text{ vs. } x_1
\]

\[
y \text{ vs. } x_2
\]

and so on.

This helps identify:

\begin{itemize}
    \item trends,
    \item bad data,
    \item outliers,
    \item nonlinear behavior.
\end{itemize}

\section{Correlations Between Variables}

Independent variables may themselves be correlated.

For example:

\[
x_2 \text{ vs. } x_3
\]

might show a strong trend.

This can produce misleading results.

\section{Correlation vs. Causation}

An important warning:

\begin{quote}
Correlation does not imply causation.
\end{quote}

Two variables may move together without one causing the other.

Regression analysis must therefore be interpreted carefully.

\section{Data Structures for Regression}

There are two common choices:

\begin{enumerate}
    \item NumPy arrays,
    \item pandas DataFrames.
\end{enumerate}

\section{Matrix Interpretation}

Suppose:

\[
y =
\beta_0
+
\beta_1 x_1
+
\beta_2 x_2
\]

for many data points.

We can write:

\[
y_1 =
\beta_0
+
\beta_1 x_{11}
+
\beta_2 x_{21}
\]

\[
y_2 =
\beta_0
+
\beta_1 x_{12}
+
\beta_2 x_{22}
\]

and so on.

\section{Matrix Form}

These equations can be rewritten as:

\[
\mathbf{y}
=
X
\mathbf{\beta}
\]

where:

\begin{itemize}
    \item \(\mathbf{y}\) is the vector of dependent values,
    \item \(X\) is the design matrix,
    \item \(\mathbf{\beta}\) is the parameter vector.
\end{itemize}

\section{Linear Algebra Perspective}

In principle:

\[
\mathbf{\beta}
=
(X^T X)^{-1} X^T \mathbf{y}
\]

This is the mathematical foundation of many regression algorithms.

\section{Important Practical Point}

For most Python regression tools:

\begin{itemize}
    \item \(X\)-data should be a two-dimensional matrix,
    \item \(y\)-data should be a one-dimensional array.
\end{itemize}

\section{Simple NumPy Example}

\begin{lstlisting}[language=Python]
import numpy as np

xi = [1, 2, 3, 4, 5]

yi = [1.1, 2.1, 2.9, 3.9, 5.1]

X = np.array(xi).reshape((-1,1))

y = np.array(yi)
\end{lstlisting}

\section{Why Reshape?}

The line:

\begin{lstlisting}[language=Python]
reshape((-1,1))
\end{lstlisting}

converts the array into a column matrix.

This is required because regression libraries expect:

\[
N \times M
\]

data structures for independent variables.

\section{Printing NumPy Arrays}

NumPy arrays display differently from Python lists.

For example:

\begin{lstlisting}[language=Python]
print(y)
\end{lstlisting}

may display:

\begin{lstlisting}
[1.1 2.1 2.9 3.9 5.1]
\end{lstlisting}

without commas.

\section{Data Cleaning}

Before fitting, we should:

\begin{itemize}
    \item plot the data,
    \item look for missing points,
    \item identify outliers,
    \item think about appropriate models.
\end{itemize}

\section{Key Takeaways}

\begin{itemize}
    \item Real regression problems often involve many variables.
    \item Regression coefficients have uncertainties.
    \item Correlations between variables can complicate interpretation.
    \item Regression is fundamentally a linear-algebra problem.
    \item Most regression libraries require two-dimensional \(X\)-data.
    \item Exploratory plotting is extremely important before fitting.
\end{itemize}

\end{document}
